---
title: "TIP-RA QC & Clustering"
author: "Virginia Muir, Kaitlin Flynn, Matt Dufort"
date: "2023-05-10"
output: html_document
---
# Analysis Summary

TIP-RA collects RA subject CD4 T cells via flow cytometry.  The goal of this analysis is likely to characterize Tmr+ cells for 5 sets of antigen-specific cells: flu, aggrecan, cilp, enolase, and vimentin/fibrinogen.  The data contain subjects at varying stages of disease, and most or all have at least one HLA-DRB1*0401 allele.

## Prepare R environment 

```{r setup, include=FALSE}
# Session options
options(stringsAsFactors = FALSE)
set.seed(12345)

library(knitr)
opts_chunk$set(echo = F, message = F, warning = F, cache = T)

# Set up the environment with required packages, installing as needed
if (!require(BiocManager)) install.packages("BiocManager")
if (!require(flowCore)) BiocManager::install("flowCore"); library(flowCore)
if (!require(flowStats)) BiocManager::install("flowStats"); library(flowStats)
if (!require(cytofkit)) {if(!require(remotes)) install.packages("remotes"); remotes::install_github("JinmiaoChenLab/cytofkit")}; library(cytofkit)
if (!require(Rphenograph)) {if(!require(remotes)) install.packages("remotes"); remotes::install_github("JinmiaoChenLab/Rphenograph")}; library(Rphenograph)
if (!require(igraph)) install.packages("igraph"); library(igraph)
if (!require(ggthemes)) install.packages("ggthemes"); library(ggthemes)
if (!require(gridExtra)) install.packages("gridExtra"); library(gridExtra)
if (!require(readxl)) install.packages("readxl"); library(readxl)
if (!require(survival)) install.packages("survival"); library(survival)
if (!require(reshape2)) install.packages("reshape2"); library(reshape2)
if (!require(tidyverse)) install.packages("tidyverse"); library(tidyverse)
if (!require(miscHelpers)) {if(!require(remotes)) install.packages("remotes"); remotes::install_github("BenaroyaResearch/miscHelpers")}; library(miscHelpers)

# Set up the ggplot default params
theme_set(
  theme_bw(12) + 
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.title = element_text(size=15, face="bold", margin = margin(10,0,10,0)),
          axis.text.x = element_text(angle=45, hjust = 1)))
```

```{r set_working_directory, cache=FALSE}
# Set working directory 
## (This isn't strictly required, since Rmarkdown will use whatever folder this script is stored in as the working directory, but I find it useful for troubleshooting in the console.)
## A windows version would look like:
## setwd("~\Grandparent folder\Parent folder\Folder where this script and your data folders are")
dir_root <-
  file.path("~", "Library", "CloudStorage", "Box-Box", "Projects",
            "TIP_RA_cross_sectional_simplified")
setwd(dir_root)
opts_knit$set(root.dir = dir_root)
```

```{r set_directories_and_global_variables}
dir_data <- "data"

# set up colorblind palette
cb_pal <- colorblind_pal()(8)[-1] 

# Create vector of markers that will be used for all downstream clustering
assay_markers = c("CD4", "CD45RA", "CD38", "CCR4", "CCR6", "CCR7", "CXCR3")
clustering_markers <- c("CD45RA", "CD38", "CCR4", "CCR6", "CCR7", "CXCR3")

# Generate file structure in which to save R data and exports
path_stem <- file.path(dir_root, "output")
dir.create(path_stem)
dir_R_objects <- file.path(path_stem, "R_objects"); dir.create(dir_R_objects)
dir_analysis_plots <- file.path(path_stem, "analysis_plots"); dir.create(dir_analysis_plots)
dir_QA_plots <- file.path(path_stem, "QA_plots"); dir.create(dir_QA_plots)
dir_density_plots <- file.path(dir_QA_plots, "density_plots"); dir.create(dir_density_plots)
```


## Load & transform data
Pull in the flow data (with dump gates removed), including sample/subject associations.

The flow data are brought in as FCS files containing MFI values. We transform MFIs using a hyperbolic arcsin transformation with b = 1/150.

```{r import_and_format_data, include=F}
## There are a couple of ways you can import your data, depending on which data you'd like to pull in.

# set data folder name
dir_data_fcs <- file.path(dir_data, "V1_FCS")

# List all files in the data folder
fcs_files_all <- list.files(dir_data_fcs, full.names = T, recursive = T)

# List all files that end in .fcs anywhere in your working directory & its sub-directories
# fcs_files_all <- list.files(pattern = ".fcs", full.names = T, recursive = T)

#Subset to total CD4 files
files_fcs_cd4_pre <- fcs_files_all[str_detect(fcs_files_all, regex("Pre[_-]Tube", ignore_case = T))]

#Subset to tmr+ files
files_fcs_flu_tmr <- fcs_files_all[str_detect(fcs_files_all, "Flu")]
files_fcs_agg_tmr <- fcs_files_all[str_detect(fcs_files_all, "Agg")]
files_fcs_cilp_tmr <- fcs_files_all[str_detect(fcs_files_all, "CILP")]
files_fcs_eno_tmr <- fcs_files_all[str_detect(fcs_files_all, "Eno")]
files_fcs_vf_tmr <- fcs_files_all[str_detect(fcs_files_all, "VF")]

# Load initial data for ID conversion
tipra_sample_info =
  read_xlsx(file.path(dir_data, "TIP_RA_sample_info.xlsx"), 
            sheet = "samples")

build_fcsList <- function(file_list, phenotypes) {
  tmp_list <- list()
  for (fcs in file_list) {
    tube_id <- fcs %>%
      str_extract("Tube_[0-9]{3}") %>%
      str_remove("Tube_")
    if(str_detect(fcs, "TIP-RA")){
      fcs_name <- phenotypes %>%
        dplyr::filter(`FCS ID` == "N/A", `Tube ID` == tube_id) %>%
        .$Labid %>%
        str_remove_all("[[:punct:]]") %>%
        paste0("lab", .)
    } else {
      fcs_name <- phenotypes %>%
        dplyr::filter(`FCS ID` != "N/A", `Tube ID` == tube_id) %>%
        .$Labid %>%
        str_remove_all("[[:punct:]]") %>%
        paste0("lab", .)
    }
    tmp_list[[fcs_name]] <- read.FCS(fcs, transformation = F)
  }
  return(tmp_list)
}

# Bring in FCS data
files_fcs_cd4_pre <- build_fcsList(files_fcs_cd4_pre, tipra_sample_info)
files_fcs_flu_tmr <- build_fcsList(files_fcs_flu_tmr, tipra_sample_info)
files_fcs_agg_tmr <- build_fcsList(files_fcs_agg_tmr, tipra_sample_info)
files_fcs_cilp_tmr <- build_fcsList(files_fcs_cilp_tmr, tipra_sample_info)
files_fcs_eno_tmr <- build_fcsList(files_fcs_eno_tmr, tipra_sample_info)
files_fcs_vf_tmr <- build_fcsList(files_fcs_vf_tmr, tipra_sample_info)

# Process data to make labels R-friendly & to specify markers of interest
key = read.csv(file.path(dir_data, "MarkerKey.csv"), header = T)

process_data <- function(fcs){
  # Tidy marker names 
  marker_key <- pData(parameters(fcs))
  pData(parameters(fcs))$desc <- key$desc[match(marker_key$name, key$name)] %>%
    toupper()

  # This changes parameters(fcs)$name, featureNames(fcs), and colnames(fcs) - aka column names for the events matrix of the fcs - all in one fell swoop.
  colnames(fcs) = pData(parameters(fcs))$desc
  
  # Preserve only the markers that will be useful for clustering - in a shared order across fcs files
  fcs = fcs[, assay_markers]
  
  # Remove boundary events (those outside of FCS's range)
  if(nrow(exprs(fcs)) > 1){
    fcs = 
      flowCore::Subset(
        fcs, 
        flowCore::filter(
          fcs, 
          boundaryFilter(x = assay_markers, 
                         tolerance = 0, 
                         side = "both")))
    }

}

# clean up data
files_fcs_cd4_pre <- map(files_fcs_cd4_pre, process_data)
files_fcs_flu_tmr <- map(files_fcs_flu_tmr, process_data)
files_fcs_agg_tmr <- map(files_fcs_agg_tmr, process_data)
files_fcs_cilp_tmr <- map(files_fcs_cilp_tmr, process_data)
files_fcs_eno_tmr <- map(files_fcs_eno_tmr, process_data)
files_fcs_vf_tmr <- map(files_fcs_vf_tmr, process_data)
```


```{r merge_and_transform_data, include=F}
# Merge total and tmr+ expression data
all_merged = files_fcs_cd4_pre
for(subj in names(files_fcs_cd4_pre)){
  exprs(all_merged[[subj]]) = 
    rbind(exprs(files_fcs_cd4_pre[[subj]]), 
          if(is.null(files_fcs_flu_tmr[[subj]]) == F) exprs(files_fcs_flu_tmr[[subj]]),
          if(is.null(files_fcs_agg_tmr[[subj]]) == F) flowCore::exprs(files_fcs_agg_tmr[[subj]]),
          if(is.null(files_fcs_cilp_tmr[[subj]]) == F) flowCore::exprs(files_fcs_cilp_tmr[[subj]]),
          if(is.null(files_fcs_eno_tmr[[subj]]) == F) flowCore::exprs(files_fcs_eno_tmr[[subj]]),
          if(is.null(files_fcs_vf_tmr[[subj]]) == F) flowCore::exprs(files_fcs_vf_tmr[[subj]]))
}

# Transform the data
asinh_tfm_data <- function(fcs){
  # Arcsinh transform remaining columns
  tl <- transformList(colnames(fcs), arcsinhTransform(a=0, b=1/150), transformationId="asinh")
  fcs = transform(fcs, tl) 
}

all_tfm <- map(all_merged, asinh_tfm_data)
fcsNames = names(all_tfm)

tfm_fs <- as(all_tfm, "flowSet")


# Extract expression data and label Tmr+ events
mergedExpr = 
  setNames(data.frame(matrix(ncol = ncol(flowCore::exprs(tfm_fs[[1]]))+2, nrow = 0)),
           c(colnames(tfm_fs), "sample", "tmr_pos"))

for (name in fcsNames) {
  tmp.expr = as.data.frame(exprs(tfm_fs[[name]]))[, assay_markers]
  tmp.expr$sample = as.character(name)
  tmp.expr$tmr_pos = 
    c(rep("none", nrow(exprs(files_fcs_cd4_pre[[name]]))), 
      if(is.null(files_fcs_flu_tmr[[name]]) == F) rep("flu", nrow(exprs(files_fcs_flu_tmr[[name]]))),
      if(is.null(files_fcs_agg_tmr[[name]]) == F) rep("agg", nrow(exprs(files_fcs_agg_tmr[[name]]))),
      if(is.null(files_fcs_cilp_tmr[[name]]) == F) rep("cilp", nrow(exprs(files_fcs_cilp_tmr[[name]]))),
      if(is.null(files_fcs_eno_tmr[[name]]) == F) rep("eno", nrow(exprs(files_fcs_eno_tmr[[name]]))),
      if(is.null(files_fcs_vf_tmr[[name]]) == F) rep("vf", nrow(exprs(files_fcs_vf_tmr[[name]]))))
  mergedExpr = rbind(mergedExpr, tmp.expr)
}


# test size of cohorts once removing, regating
RA_tmrs <- c("agg", "cilp", "eno", "vf")

test_size <- mergedExpr %>% dplyr::filter(tmr_pos %in% RA_tmrs)
```    


```{r phenograph_clustering_functions_for_discovr}
# Clustering markers object is defined in the first set-up chunk.  Tweak which markers are included up there.

# Set up Phenograph function to use kd tree
find_neighbors <- function(data, k){
  nearest <- RANN::nn2(data, data, k, treetype = "kd", searchtype = "standard")
  return(nearest[[1]])
}

Rpheno <- function(data, k=30){
  if(is.data.frame(data))
    data <- as.matrix(data)

  if(!is.matrix(data))
    stop("Wrong input data, should be a data frame or matrix!")

  if(k<1){
    stop("k must be a positive integer!")
  }else if (k > nrow(data)-2){
    stop("k must be smaller than the total number of points!")
  }

  message("Run Rphenograph starts:","\n",
          "  -Input data of ", nrow(data)," rows and ", ncol(data), " columns","\n",
          "  -k is set to ", k)

  cat("  Finding nearest neighbors...")
  t1 <- system.time(neighborMatrix <- find_neighbors(data, k=k+1)[,-1])
  cat("DONE ~",t1[3],"s\n", " Compute jaccard coefficient between nearest-neighbor sets...")
  t2 <- system.time(links <- Rphenograph:::jaccard_coeff(neighborMatrix))

  cat("DONE ~",t2[3],"s\n", " Build undirected graph from the weighted links...")
  links <- links[links[,1]>0, ]
  relations <- as.data.frame(links)
  colnames(relations)<- c("from","to","weight")
  t3 <- system.time(g <- igraph::graph.data.frame(relations, directed=FALSE))

  cat("DONE ~",t3[3],"s\n", " Run louvain clustering on the graph ...")
  t4 <- system.time(community <- igraph::cluster_louvain(g))
  cat("DONE ~",t4[3],"s\n")

  message("Run Rphenograph DONE, took a total of ", sum(c(t1[3],t2[3],t3[3],t4[3])), "s.")
  cat("  Return a community class\n  -Modularity value:", igraph::modularity(community),"\n")
  cat("  -Number of clusters:", length(unique(igraph::membership(community))),"\n")

  return(community)
}


# Run phenograph (using kd treetype) on each subject.
PhenographClust = function(fcs, clustering_markers) {
  exprs_mat = as.matrix(as.data.frame(flowCore::exprs(fcs))[,clustering_markers])
  RPvect = as.numeric(igraph::membership(Rpheno(data = exprs_mat)))
  return(RPvect)
}

```

```{r phenograph_clustering_and_cluster_calculations}
# check for cached version of clustering results (for time and stochasticity)
pheno_filename = file.path(dir_R_objects, "all_phenograph_data.RData")

if (file.exists(pheno_filename)) {
  load(pheno_filename)
} else {
  # run clustering and store it
  mergedExpr$RPclust = unlist(lapply(all_tfm, PhenographClust, clustering_markers))
  
  # Get summary of the number of clusters generated for each subject
  n_pheno_clusts <- mergedExpr %>%
    group_by(sample) %>%
    dplyr::summarize(k_clusters = max(RPclust))
  # View(n_pheno_clusts) # 10 - 24 clusters per subject, mostly in the 16 - 19 range
  
  ## Calculate means by cluster
  
  # Calculate mean expression value of each marker for each phenograph cluster in each subject
  RP_mean <- mergedExpr %>%
    dplyr::select(-tmr_pos, -contains("tsne"), -contains("umap")) %>%
    group_by(sample, RPclust) %>%
    summarise_all(mean) %>%
    mutate(RPclust = as.character(RPclust))
  
  # Calculate total CD8 mean expression for each subject
  CD4_mean = mergedExpr %>%
    dplyr::select(-tmr_pos, -RPclust) %>%
    group_by(sample) %>%
    summarise_all(mean) %>%
    mutate(RPclust = "Total_CD4")
  
  RP_mean = bind_rows(RP_mean, CD4_mean)
  
  # Count cells of each specificity in each phenograph cluster (from each sample)
  RPtmr_counting = mergedExpr %>% 
    dplyr::select(sample, tmr_pos, RPclust) %>%
    group_by(sample, RPclust) %>%
    dplyr::summarise(
      clust_size = n(), 
      none = sum(tmr_pos=="none"), 
      flu = sum(tmr_pos=="flu"),
      agg = sum(tmr_pos=="agg"), 
      cilp = sum(tmr_pos=="cilp"), 
      eno = sum(tmr_pos=="eno"),
      vf = sum(tmr_pos=="vf"))
  
  aggregate_counts = RPtmr_counting %>%
    dplyr::select(-RPclust) %>%
    group_by(sample) %>%
    summarise_all(sum) %>%
    rename_at(vars(-sample),function(name) paste0(name,"_tot"))
  
  RPtmr_counting = RPtmr_counting %>%
    left_join(aggregate_counts) %>%
    mutate(pct_nonspec_in_clust = none/none_tot * 100,
           pct_flu_in_clust = flu/flu_tot * 100,
           pct_agg_in_clust = agg/agg_tot * 100,
           pct_cilp_in_clust = cilp/cilp_tot * 100,
           pct_eno_in_clust = eno/eno_tot * 100,
           pct_vf_in_clust = vf/vf_tot * 100)
  
  # Save the clustering output
  save(mergedExpr, clustering_markers, RP_mean, RPtmr_counting, 
       file = pheno_filename)
}
```


```{r downsample_for_qa}
# Check for donors with too many/too few collected events
events_by_donor = mergedExpr %>%
  group_by(sample) %>%
  dplyr::summarise(n_events = n())
hist(events_by_donor$n_events)

# Make a df with the same total number of total CD4 events for each subject 
all_events_ds <- mergedExpr %>% 
  dplyr::filter(sample %in% events_by_donor$sample[events_by_donor$n_events > 4500],
                tmr_pos == "none") %>%
  group_by(sample) %>% 
  sample_n(4500) %>%
  bind_rows(mergedExpr %>% 
              dplyr::filter(sample %in% events_by_donor$sample[events_by_donor$n_events > 4500],
                            tmr_pos != "none"))

# Make a separate downsampled version of all_tmf
# Check for donors with too many/too few collected events

# Downsample to a fixed amount of cells for faster processing/unbiased clustering
ds_events = function(fcs) {
  flowCore::exprs(fcs) = flowCore::exprs(fcs) %>%
    .[sample(nrow(.), 4500, replace = F), ]
  return(fcs)
}

ds_fs <- all_tfm %>%
  .[events_by_donor$sample[events_by_donor$n_events > 4500]] %>%
  map(ds_events) %>%
  as("flowSet")
```


## QA, Density plots by sample
Check marker density plots for each sample to look for any major batch effects.  Several markers look quite variable

```{r density_plots_by_sample_qa, message=F, fig.width = 9, fig.height = 12}

pdf(file = file.path(dir_density_plots, "density_by_sample.pdf"), width = 9, height = 12)
flowViz::densityplot(
  name~., ds_fs, channels=assay_markers, 
  groups=name, main="Post-Transformation", 
  scales=list(y=list(draw=T)), 
  filter=lapply(assay_markers, curv1Filter),
  layout=c(7,1))
invisible(dev.off())
```

